---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# zerozarr

<!-- badges: start -->
<!-- badges: end -->


Here we cut a 3D array into 12 pieces, as a start at illustrating how Zarr works. 

Note that everything here will be wrong!! For a little while, I need to step forward and backwards and look at everything carefully, but I want it visible. Reach out if you're interested, I don't really think I can write a Zarr implementation, but I can write a toy version with enough chops to show what's happening to folks who know R. 

For now I'm sticking with 3D only because it's easy to think about and show easy 
illustrations, little chunks of RGB array so I know I have something sensible and not just one of those sliding tile games. 

You can watch me slowly increment this to something useful, I'm obsessed with how cool Zarr is how it is composed of very general pieces. (And, I know about {Rarr} and {pizarr} and GDAL multidim and stars and terra, not wanting to undercut those this is just a learning exercise for now). 


## TODO

- n-d tiling (generalize {grout})
- create json metadata
- create parquet metadata
- make an example that avoids degenerate coordinate referencing ...
- encoding of chunks and read/write (we have write-via-GDAL and read via r-lib/archive, but only with NONE or GZIP compression)


Also maybe we can do the kerchunk referencing thing with some example NetCDF files. 



```{r}
pngfile <- system.file("img", "Rlogo.png", package="png", mustWork = TRUE)

## we have this massive array
a <- fastpng::read_png(pngfile)
dim(a)

## let's transpose and flip
a <- aperm(a, c(2, 1, 3))
## we have a tiling logic, let's go for 26x25x4

## hypertidy/grout on github
tiling <- grout::grout(dim(a)[1:2], blocksize = c(25, 26))
idx <- grout::tile_index(tiling)
idx

## we necessarily have some dangle, but we minimized it by choosing 26 not 25 for x
plot(tiling)

subset_3d_array <- function(x, offset, size) {
  ## offsets are zero-based
  l <- apply(rbind(offset + 1, size), 2, \(.x) seq(.x[1], length.out = .x[2]))
  x[l[[1]], l[[2]], l[[3]]]
}
## offset and size are the values in idx
str(subset_3d_array(a, c(0, 0, 0), c(26, 25, 4)))
## our table idx describes the blocks, and their sizes (some are 26x25x4, some are 24x25x4)
split_array_blocks <- function(x, index, n3 = 4) {
  index$offset_z <- 0
  index$nz <- n3
  offsets <- lapply(purrr::transpose(index[c("offset_x", "offset_y", "offset_z")]), unlist)
  sizes <- lapply(purrr::transpose(index[c("ncol", "nrow", "nz")]), unlist)
  
  l <- vector("list", nrow(index))
  for (i in seq_along(offsets)) {
    l[[i]] <- subset_3d_array(x, offsets[[i]], sizes[[i]])
  }
  l
}

## this is the list of 0.0.0 0.1.0 1.0.0 1.1.0 ... 2.3.0 
## blocks, there are 3 tile columns, and 4 tile rows, and 1 tile in z
## so we go up as far as 2.3.0
listarr <- split_array_blocks(a, idx)

## ok so we're not n-dimension yet
chunk <- apply(cbind(0, idx$tile_row - 1, idx$tile_col - 1), 1, paste0, collapse = ".")
```


That indexing for the chunks 0.0.0 --> 2.3.0 will hold the implicit arrangement of the tiles. 

We also have an xmin/xmax/ymin/ymax for each chunk, which is extremely convenient. 

```{r plot}
mar <- c(2.8, 2.5, 0, 0)
cex.axis <- .5
op <- par(bg = "grey35", mar = mar, cex.axis = cex.axis)
plot(NA, xlab = "", ylab = "", asp = 1,
     xlim = c(0, max(idx$xmax)), ylim = c(0, max(idx$ymax)))
for (i in seq_along(listarr)) {
  ex <- unlist(idx[i, c("xmin", "xmax", "ymin", "ymax")])
  ximage::ximage(aperm(listarr[[i]], c(2, 1, 3)), ex, add = TRUE)
  p <- c(mean(ex[1:2]), mean(ex[3:4]))
  vaster::plot_extent(ex, add = TRUE)
  text(p[1], p[2], lab = chunk[i], cex = 2, col = "hotpink")

}
par(op)
```

Now, let's create an actual ZARR in version 3 with GDAL, each block has GZIP compression. We roundtrip our PNG file into a Zarr store, and plot in the same way as we did with our manually split array above. 

```{r realzarr}
Rlogo.zarr <- file.path(tempdir(), "Rlogo.zarr")

unlink(Rlogo.zarr, recursive = TRUE)

block <- c(25, 26, 4)
blk <- paste0(rev(block), collapse= ",")


## COMPRESS = NONE/BLOSC/ZLIB/GZIP/LZMA/ZSTD/LZ4]  (no ZLIB, LZMA, ZSTD, LZ4 with ZARR_V3)
## filter ‘none’, ‘gzip’, ‘bzip2’, ‘compress’, ‘lzma’, ‘xz’, ‘uuencode’, ‘lzip’, ‘lrzip’, ‘lzop’, ‘grzip’, ‘lz4’, ‘zstd’.
system(glue::glue("
       gdal_translate {pngfile} {Rlogo.zarr} -of ZARR -co INTERLEAVE=BAND  -co COMPRESS=GZIP -co BLOCKSIZE={blk} -co FORMAT=ZARR_V3
"))



files <- fs::dir_ls(Rlogo.zarr,  recurse = TRUE, regexp = "Rlogo.zarr/Rlogo/c/[0-9]", type = "f")
files

op <- par(bg = "grey35", mar = mar, cex.axis = cex.axis)
## setup out plot as before
plot(NA, xlab = "", ylab = "", asp = 1,
     xlim = c(0, max(idx$xmax)), ylim = c(0, max(idx$ymax)))


## in this loop, 'ar' is now identical to the corresponding element in 
for (i in seq_along(files)) {
  ar <- readBin(archive::file_read(files[i], mode = "rb"), "integer", size = 1, n = prod(block), signed = FALSE)
  ex <- unlist(idx[i, c("xmin", "xmax", "ymin", "ymax")])
  ximage::ximage(aperm(array(ar, block), c(2, 1, 3)), ex, add = TRUE)
  p <- c(mean(ex[1:2]), mean(ex[3:4]))
  vaster::plot_extent(ex, add = TRUE)
  
  text(p[1], p[2], lab = gsub(file.path(dirname(Rlogo.zarr), "Rlogo.zarr"), ".", files[i]), cex = 1, col = "hotpink")
  
}
par(op)
```


## Code of Conduct
  
Please note that the zerozarr project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
